\begin{thebibliography}{52}
\bibitem{akopyan} Akopyan, A. V.,
\textit{A piecewise linear analogue of Nash--Kuiper theorem},
a preliminary version (in Russian) can be found on
\href{http://www.moebiuscontest.ru/files/2007/akopyan.pdf}{www.moebiuscontest.ru}

\bibitem{alexander} Alexander, R. 
\textit{Lipschitzian mappings and total mean curvature of polyhedral surfaces,} 
no. I, Trans. Amer. Math. Soc. 288 (1985), no. 2, 661--678.


\bibitem{alexandrov} Alexandrov, A. D. \textit{Convex Polyhedra.} Springer Monographs in Mathematics 2005

\bibitem{bezdek-connelly} Bezdek, K.;  Connelly, R. 
\textit{Pushing disks apart---the Kneser--Poulsen conjecture in the plane.}  
J. Reine Angew. Math.  553  (2002), 221--236.

\bibitem{blaschke} Blaschke, W., \textit{Kreis und Kugel} ???
???Russian translation ??? 

\bibitem{burago-zalgaller}
Burago, Yu. D. and Zalgaller, V. A.: Isometric piecewise linear immersions of two-dimensional manifolds with polyhedral metrics into $\RR^3$
, St. Petersburg Math. J., Vol.7, No. 3, 369--385, 1996.

\bibitem{brehm} Brehm, U., \textit{Extensions of distance reducing mappings to piecewise congruent mappings on $R\sp{m}$.}  J. Geom.  16  (1981), no. 2, 187--193.

\bibitem{BBI}  Burago, D.;  Burago, Yu.;  Ivanov, S. \textit{A course in metric geometry}.

\bibitem{cauchy} A. Cauchy, Sur les polygones et les poly\`{e}dres (Second M\'emoire, Oeuvres  Compl`{e}tes, II$^\mathrm{e}$
S\'erie, vol. 1; Paris, 1905).

\bibitem{connelly-flex}
Connelly, Robert, A counterexample to the rigidity conjecture for polyhedra, 
Publications Math\'ematiques de l'IH\'ES 47, (1977), 333--338,


\bibitem{TF} D. Fuchs and S. Tabachnikov \textit{Mathematical omnibus. Thirty lectures on classic mathematics} 

\bibitem{frolik} Z. Frol\'{\i}k, \textit{Concerning topological convergence of sets}, Czechoskovak Math. J. 10 (1960), 168--180

\bibitem{gromov-polynomial-growth}M. Gromov, Groups of Polynomial growth and Expanding Maps, Publications Math\'ematiques de l'IH\'ES 53, 1981

\bibitem{hausdorff} Hausdorff, F. \textit{Grundz\"uge der Mengenlehre} 1914

\bibitem{eliashberg-mishachev} Eliashberg, Y.  Mishachev, N. Introduction to the \textit{h}-Principle American Mathematical Soc., 2002

\bibitem{krat} Krat, S. \textit{Approximation Problems in Length Geometry,} Thesis, 2005

\bibitem{sabitov} Sabitov, I. Kh, On the problem of the invariance of the volume of a deformable polyhedron,  
Uspekhi Matematicheskikh Nauk 50 (2): 223--224

%\bibitem{pak} Pak, I. \href{http://www.math.ucla.edu/~pak/book.htm}{\textit{Lectures on Discrete and Polyhedral Geometry}}

\bibitem{petrunin-ruble}  Petrunin, A.  Arnold's problem on paper folding (Russian), \textit{Problems of S.Petersburg's olimpiads in mathematics 2008}, pp. 116--125; \textit{arXiv:1004.0545}

\bibitem{petrunin-inverse} Petrunin, A.
On intrinsic isometries to Euclidean space, 
\textit{St. Petersburg Mathematical Journal}, 22 (2011), 803--812.

\bibitem{rudin}Rudin, W. Principles of Mathematical Analysis

\bibitem{saraf}  Saraf, S. Acute and nonobtuse triangulations of polyhedral surfaces ??? 

\bibitem{schoenberg-zaremba}
Schoenberg, I. J.; Zaremba, S. C.
On Cauchy's lemma concerning convex polygons. 
Canad. J. Math. 19 1967 1062--1071.

\bibitem{wijsman} Wijsman, Robert A. Convergence of sequences of convex sets, cones and functions. II. Trans. Amer. Math. Soc.  123 (1) 1966, 32--45

\bibitem{zalgaller}Zalgaller, V. A.
\textit{On deformations of a polygon on a sphere.} (Russian)
Uspehi Mat. Nauk (N.S.) 11 (1956), no. 5(71), 177--178.

\bibitem{zalgaller-PL} Zalgaller, V. A. 
\textit{Isometric imbedding of polyhedra.} (Russian)  Dokl. Akad. Nauk SSSR  123  1958 599--601.
\end{thebibliography}
